Semantic Preserving Bijective Mappings for Expressions involving Special Functions between Computer Algebra Systems and Document Preparation Systems

### 2018-11-30 main.tex
Try to avoid informal words (addresses #29)

\section{Introduction}
A typical workflow of a scientist who writes a scientific publication is to use \gls*{dps} to write the paper and one or more \gls*{cas} for verification, analysis and visualization. Especially in the \gls*{stem} literature, \LaTeX{}
%\footnote{Note that technically \LaTeX{} is not a \gls*{dps} (\url{https://www.latex-project.org/about/}, seen 07/2017) like \textit{Microsoft Word}. However, since \LaTeX{} (an extension of \TeX) is the de facto standard for writing articles in \gls*{stem}, and we only focus on mathematical writing in this paper, we have categorized \LaTeX{} as a \gls*{dps} as well.} 
has become the de facto standard for writing scientific publications over the past 30 years~\parencites{Knuth}[559]{DigitalTypo}{LATEX:Standard}. \LaTeX{} enables printing of mathematical formulae in a structure similar to handwritten style. For example, consider the specific Jacobi polynomial~\parencite[Table 18.3.1]{NIST:DLMF}
\begin{equation}\label{eq:P}
P_n^{(\alpha , \beta)}(\cos(a\Theta)),
\end{equation}
where $n$ is a nonnegative integer, $\alpha, \beta > -1$, and $a,\Theta\in \Real$.
This mathematical expression can be written in \LaTeX{} as
\begin{equation*}
\verb|P_n^{(\alpha,\beta)}(\cos(a\Theta))|.
\end{equation*}

@@

Another more questionable translation decision is alphanumerical expressions. As explained in~\Cref{tab:allTypesTable}, the \gls*{pom}-tagger handles strings of letters and numbers differently depending on the order of the symbols. The reason is that an expression such as `$4b$' is usually considered to be a multiplication of $4$ and `$b$,' while `$b4$' 
displays like indexing `$b$' by $4$. While the first example produces two nodes, namely $4$ and `$b$', the second example `$b4$' produces just a single alphanumerical node in the \gls*{pom-pt}. The translator interprets alphanumerical expressions as multiplications for two reasons:~(1) we would assume that the inputs `$4b$' and `$b4$' are mathematically equivalent; and (2) it is more common in mathematics to use single letter names for variables~\parencite{Notation:History}. Therefore we have used rules as follows {\bf [HSC: I changed the wording here, definition$\to$rules. Is this ok? Otherwise further clarify.]}

@@

\section{The Translator}
The translator analyzes a parse tree to perform translations. For generating a parse tree of \LaTeX{} expressions, the translator uses the \gls*{pom}-Tagger~\parencite{POM-Tagger}\footnote{Named according to the Part-of-Speech-Taggers in \gls*{nlp}.}. \gls*{cas} define their own syntax parser. We were able to use \Maple's internal data structure to 
obtain {\bf [HSC: I changed get$\to$obtain.]} a parse tree of the input. \Cref{sec:forward-translation} and \Cref{sec:backward-translation} will explain the parsing and translation process in detail.

@@

The parsed tree generated by the \gls*{pom}-Tagger is not a mathematical expression tree. The \gls*{pom} project aims to disambiguate mathematical \LaTeX{} expressions and generates an expression tree. However, in the current state, many expressions cannot yet be disambiguated. Therefore, the \gls*{pom}-tagger generates a raw parsed tree where each token in the \LaTeX{} expression is a node in the tree. We call this parsed tree the \gls*{pom-pt}.

@@

With these tools, we can translate a \LaTeX{} expression by translating the \gls*{pom-pt} node by node and perform group or reordering operations for some special cases. The algorithm is realized in a simple recursive structure. Whenever the algorithm finds a leaf, it can translate this single term. If the node is not a leaf, it starts to translate all children of the node recursively. This idea seems to be a practical and elegant solution. However, it has a significant drawback. It cannot be used to translate functions. Since the arguments of functions are following siblings in the \gls*{pom-pt}, the algorithm needs to look ahead when a leaf is a known function, e.g., in the case of a semantic macro with arguments (see~\Cref{fig:syntax-tree-usecase}). Algorithm~\ref{alg:translation} is an improved version with look ahead functionality.

@@

\subsection{Problems with the Look Ahead Approach}
The look ahead functionality seems to solve the problems for functions. But there is another problem that was not mentioned in \Cref{sec:problems}, because it seems to be not a problem previously. In some cases the arguments of a function do not follow but precede the function node.

@@

If we take a closer look at mathematical notations we discover many different types of notations used to represent formulae. \Cref{tab:notations} illustrates the expression $(a+b)x$ in different notations. The \gls*{pol}\footnote{Also known as \textit{prefix notation}, \textit{Warsaw Notation} or \textit{{\L}ukasiewicz notation}. It was invented by J. {\L}ukasiewicz 1924 to create a parenthesis-free notation~\parencite{Hamblin1962}. Note that this notation is indeed parenthesis-free as long as all operators have the same arity.} (hereafter called prefix notation) places the operator to the left of/before its operands. The \gls*{rpol}\footnote{Also known as \textit{postfix notation}. Also invented by J. {\L}ukasiewicz. Same as \gls*{pol} it does not need parenthesis as long as all operators have the same arity.} (hereafter called postfix notation) does the opposite and places the operator to the right of/after its operands. The infix notation is commonly used in arithmetic and places the operator between its operands. This only makes sense if the operator is a binary operator.

@@

Algorithm~\ref{alg:translation} does not allow this translation right now. It has no access to previously translated nodes in its current state. This problem is solved by the \gls*{teo} that stores and groups translated objects like lists. This allows one to access the latest translated expression and use it as the argument for the factorial function. Table~\ref{tab:teo-list} shows three examples for the \gls*{teo} list that groups some tokens.

@@

{\bf [HSC: This is a new paragraph, but it does not have standard vertical space above it. I do not know why this is. Can you fix this?]} The translation process for the \Macro s is complex, so there is a special class, the \verb|MacroTranslator|, that handles those nodes in the \gls*{pom-pt}. Algorithm~\ref{alg:macro-translation} explains the \verb|MacroTranslator| without error handling. It has extracted necessary information from the \gls*{pom-pt}, such as how many arguments this function has, in Line~\ref{line:get_info}. It also processes the following siblings to translate the arguments. The \verb|MacroTranslator| will be called in Line~\ref{line:transLeaf} in Algorithm~\ref{alg:translation}, since the macro is a leaf node in the \gls*{pom-pt}. The following cases describe the different kinds of the following siblings after a semantic macro node. Those can be:
\begin{itemize}
\item an exponent, such as for `\verb|^2|' right after the macro node (Line~\ref{line:next_caret});
\item an optional parameter in square brackets right after the macro node or after an exponent (Line~\ref{line:next_optional});
\item a parameter in curly brackets (a \textit{sequence} node in the \gls*{pom-pt}) if none of the above and no `$@$' symbols were passed yet (Line~\ref{line:trans_paras});
\item `$@$' symbols (Line~\ref{line:skip_ats}); or
\item a variable in curly brackets (a \textit{sequence} node) after the `$@$' symbols were 
passed (Line~\ref{line:trans_vars}).
\end{itemize}

@@

\section{\Maple{} to Semantic \LaTeX{} Translator}\label{sec:backward-translation}
In this section, we will discuss several techniques to get access to the parse tree of \Maple's input. The translation process from this parse tree then follows the same principle as for the forward translations. Instead of writing a custom \Maple{} syntax parser, we use \Maple's internal data structure to get a syntax tree of the input\footnote{A license of \Maple{} is mandatory to perform backward translations. Our translator uses the version \Maple{} 2016 {\bf [HSC: I reworded this. Is it ok? Otherwise fix.]}.}. \Maple{} allows several different input styles. The \texttt{1D} input is mainly used for programming purposes 
and is also used to perform our translations. Internally, \Maple{} uses a \gls*{dag} for syntax trees.

@@

\subsection{Automatic Changes of Inputs in Maple}\label{subsec:maple-probs}
\Maple{} evaluates inputs automatically and changes the input into an internal representation. This internal representation might look {\bf [HSC: can you find a better word than `look'?]} a bit different to the input. One example has already been given with \Cref{fig:internal-maple-dag}, where each summand of a sum is stored with a factor. Here is a list of all internal changes that occur for inputs.

@@

The tasks for future work are diverse. The main task is to improve the translator by implementing more functions and features. For example, for the current state, only translations to \Maple's standard function library were implemented. \Maple{} allows one to load extra packages dynamically and therefore support an enhanced set of functions. This feature would drastically increase the number of possible translations. With such improvements, further work on evaluation techniques become worthwhile to evaluate the \gls*{dlmf} and \gls*{cas}.
Increasing the amount of translatable formulae in the \gls*{dlmf} and improving the verificiation 
techniques are also parts of ongoing projects {\bf [HSC: I moved this sentence from the end of the section to the end of this paragraph. Is this ok?]}.

>>>>

\section{Introduction}
A typical workflow of a scientist who writes a scientific publication is to use \gls*{dps} to write the paper and one or more \gls*{cas} for verification, analysis and visualization. Especially in the \gls*{stem} literature, \LaTeX{}
has become the de facto standard for writing scientific publications over the past 30 years~\parencites{Knuth}[559]{DigitalTypo}{LATEX:Standard}. \LaTeX{} enables printing of mathematical formulae in a structure similar to handwritten style. For example, consider the specific Jacobi polynomial~\parencite[Table 18.3.1]{NIST:DLMF}
\begin{equation}\label{eq:P}
P_n^{(\alpha , \beta)}(\cos(a\Theta)),
\end{equation}
where $n$ is a nonnegative integer, $\alpha, \beta > -1$, and $a,\Theta\in \Real$.
This mathematical expression can be written in \LaTeX{} as
\begin{equation*}
\verb|P_n^{(\alpha,\beta)}(\cos(a\Theta))|.
\end{equation*}

@@

Another more questionable translation decision is alphanumerical expressions. As explained in~\Cref{tab:allTypesTable}, the \gls*{pom}-tagger handles strings of letters and numbers differently depending on the order of the symbols. The reason is that an expression such as `$4b$' is usually considered to be a multiplication of $4$ and `$b$,' while `$b4$' gives the impression that $4$ indexing `$b$'. While the first example produces two nodes, namely $4$ and `$b$', the second example `$b4$' produces just a single alphanumerical node in the \gls*{pom-pt}. The translator interprets alphanumerical expressions as multiplications for two reasons:~(1) we would assume that the inputs `$4b$' and `$b4$' are mathematically equivalent; and (2) it is more common in mathematics to use single letter names for variables~\parencite{Notation:History}. Therefore we have used rules as follows {\bf [HSC: I changed the wording here, definition$\to$rules. Is this ok? Otherwise further clarify.]}

@@

\section{The Translator}
The translator analyzes a parse tree to perform translations. For generating a parse tree of \LaTeX{} expressions, the translator uses the \gls*{pom}-Tagger~\parencite{POM-Tagger}\footnote{Named according to the Part-of-Speech-Taggers in \gls*{nlp}.}. \gls*{cas} define their own syntax parser. We were able to use \Maple's internal data structure to 
obtain a parse tree of the input. \Cref{sec:forward-translation} and \Cref{sec:backward-translation} will explain the parsing and translation process in detail.

@@

The parsed tree generated by the \gls*{pom}-Tagger is not a mathematical expression tree. The \gls*{pom} project aims to disambiguate mathematical \LaTeX{} expressions and generates an expression tree. However, in the current state, many expressions still cannot be disambiguated. Therefore, the \gls*{pom}-tagger generates a raw parsed tree where each token in the \LaTeX{} expression is a node in the tree. We call this parsed tree the \gls*{pom-pt}.

@@

With these tools, we can translate a \LaTeX{} expression by translating the \gls*{pom-pt} node by node and perform group or reordering operations for some special cases. The algorithm is realized in a simple recursive structure. Whenever the algorithm finds a leaf, it can translate this single term. If the node is not a leaf, it starts to translate all children of the node recursively. This idea appears to be a practical and elegant solution. However, it has a significant drawback. It cannot be used to translate functions. Since the arguments of functions are following siblings in the \gls*{pom-pt}, the algorithm needs to lookahead when a leaf is a known function, e.g., in the case of a semantic macro with arguments (see~\Cref{fig:syntax-tree-usecase}). Algorithm~\ref{alg:translation} is an improved version with lookahead functionality.

@@

\subsection{Problems with the Lookahead Approach}
The lookahead functionality appears to solve the problems for functions. However, there is a problem with the lookahead functionality that \Cref{sec:problems} did not address. In some cases, the arguments of a function do not follow but precede the function node.

@@

If we intently examine mathematical notations, we discover many different types of notations used to represent formulae. \Cref{tab:notations} illustrates the expression $(a+b)x$ in different notations. The \gls*{pol}\footnote{Also known as \textit{prefix notation}, \textit{Warsaw Notation} or \textit{{\L}ukasiewicz notation}. It was invented by J. {\L}ukasiewicz 1924 to create a parenthesis-free notation~\parencite{Hamblin1962}. Note that this notation is indeed parenthesis-free as long as all operators have the same arity.} (hereafter called prefix notation) places the operator to the left of/before its operands. The \gls*{rpol}\footnote{Also known as \textit{postfix notation}. Also invented by J. {\L}ukasiewicz. Same as \gls*{pol} it does not need parenthesis as long as all operators have the same arity.} (hereafter called postfix notation) does the opposite and places the operator to the right of/after its operands. The infix notation is commonly used in arithmetic and places the operator between its operands. This only makes sense if the operator is a binary operator.

@@

Algorithm~\ref{alg:translation} does not allow this translation right now. It has no access to previously translated nodes in its current state. This problem is solved by the \gls*{teo} that stores and groups translated objects as lists. This allows one to access the latest translated expression and use it as the argument for the factorial function. Table~\ref{tab:teo-list} shows three examples for the \gls*{teo} list that groups some tokens.

@@

{\bf [HSC: This is a new paragraph, but it does not have standard vertical space above it. I do not know why this is. Can you fix this?]} The translation process for the \Macro s is complex, so there is a special class, the \verb|MacroTranslator|, that handles those nodes in the \gls*{pom-pt}. Algorithm~\ref{alg:macro-translation} explains the \verb|MacroTranslator| without error handling. It has extracted necessary information from the \gls*{pom-pt}, such as how many arguments this function has, in Line~\ref{line:get_info}. It also processes the following siblings to translate the arguments. The \verb|MacroTranslator| will be called in Line~\ref{line:transLeaf} in Algorithm~\ref{alg:translation}, since the macro is a leaf node in the \gls*{pom-pt}. The following cases describe the different kinds of the following siblings after a semantic macro node. Those can be:
\begin{itemize}
\item an exponent, such as for `\verb|^2|' right after the macro node (Line~\ref{line:next_caret});
\item an optional parameter in square brackets right after the macro node or after an exponent (Line~\ref{line:next_optional});
\item a parameter in curly brackets (a \textit{sequence} node in the \gls*{pom-pt}) if none of the above and no `$@$' symbols were passed (Line~\ref{line:trans_paras});
\item `$@$' symbols (Line~\ref{line:skip_ats}); or
\item a variable in curly brackets (a \textit{sequence} node) after the `$@$' symbols were 
passed (Line~\ref{line:trans_vars}).
\end{itemize}

@@

\section{\Maple{} to Semantic \LaTeX{} Translator}\label{sec:backward-translation}
In this section, we will discuss several techniques to access the parse tree of \Maple's input. The translation process from this parse tree then follows the same principle as for the forward translations. Instead of writing a custom \Maple{} syntax parser, we use \Maple's internal data structure to obtain the syntax tree of the input\footnote{A license of \Maple{} is mandatory to perform backward translations. Our translator uses the version \Maple{} 2016 {\bf [HSC: I reworded this. Is it ok? Otherwise fix.]}.}. \Maple{} allows several different input styles. The \texttt{1D} input is mainly used for programming purposes 
and is also used to perform our translations. Internally, \Maple{} uses a \gls*{dag} for syntax trees.

@@

\subsection{Automatic Changes of Inputs in Maple}\label{subsec:maple-probs}
\Maple{} evaluates inputs automatically and changes the input into an internal representation. This internal representation may differ to the input. One example has already been given with \Cref{fig:internal-maple-dag}, where each summand of a sum is stored with a factor. Here is a list of all internal changes that occur for inputs.

@@

The tasks for future work are diverse. The main task is to improve the translator by implementing more functions and features. For example, for the current state, only translations to \Maple's standard function library were implemented. \Maple{} allows one to load extra packages dynamically and therefore support an enhanced set of functions. This feature would drastically increase the number of possible translations. With such improvements, further work on evaluation techniques become worthwhile to evaluate the \gls*{dlmf} and \gls*{cas}.
Increasing the amount of translatable formulae in the \gls*{dlmf} and improving the verificiation 
techniques are also parts of ongoing projects.
